Social inequality in mortality among adults and elderly in northern Sweden 1851--2013\footnote{Paper presented at the session 'Health and Mortality in Arctic Populations'. The Ninth International Congress on Arctic Social Sciences, Umeå, June 8--12, 2017}

### 2017-06-01 icass.Rnw
Adding comments on tables

\bibliographystyle{apalike}
%\bibliographystyle{plainnat}

@@

The analyses are performed in the \R\ environment for statistical computing
and graphics \citep{cran}, especially using the package {\tt eha} 
\citep{gb12,eha}
The proportional hazards model allow us to scrutinise the differences and
the development in more detail. We divide the analysis into two age groups,
40--64 and 65--89 years of age, the first age group representing people that
were still mainly in the workforce, and the second those that mainly were 
retired. The model controls for marital
status and whether the individual resides in the urban or rural part. There
are no serious signs of deviation from proportionality assumption, except
regarding {\sc Hisclass}. Each period are analysed separately, but we then combine results
in order to show the development of the social patterning of mortality
during the studied period.

@@

The important explanatory variable, {\sc Hisclass}, is included in the
proportional hazards models as a \emph{stratification} variable. The main
reason for this choice is that the effect of {\sc Hisclass} on survival is
non-proportional, that is, it varies with age. With the reduced data set
and Poisson regression, this is simply achieved by introducing an
interaction term between {\tt age} and {\tt hisclass}. Then, the main
results are presented \emph{graphically}.

@@

Models are fitted separately for each time period, from  1851 to
2013, with the exception of the period 1951--1960 and separately for sex. We stratify on HISCLASS
and control for the covariates {\tt   civst} and urban/rural residence. The cumulative
hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
main target in our analysis: These numbers are used in the process of
graphically illustrating the development over time.

@@

Performing the same types of analyses as for all-cause mortality, we start by looking at cumulative hazards by {\sc Hisclass} at age 65, given survival to age 40. They are adjusted for {\tt marital status} and urban/rural residence. We present separate analyses for women and men.

@@

In this section we focus on cardiovascular mortality among those that mostly have left the workforce, the age group 65--89 years. In accordance with the analyses for the age group 40--64, the model adjusts for {\tt civil status} and urban/rural residence and with separate analyses for women and men. We use as a summary measure the value of the cumulative hazards at
duration 25 (age 90), starting from age 65.

>>>>

%\bibliographystyle{apalike}
\bibliographystyle{plainnat}

@@

The analyses are performed in the \R\ environment for statistical computing
and graphics \citep{cran}, especially using the package {\tt eha} 
\citep{gb12,eha}
The proportional hazards model allow us to scrutinise the differences and
the development in more detail. We divide the analysis into two age groups,
40--64 and 65--89 years of age, the first age group representing people that
were still mainly in the workforce, and the second those that mainly were 
retired. The model controls for marital
status and whether the individual resides in the urban or rural parts of the regions. There
are no serious signs of deviation from proportionality assumption, except
regarding {\sc Hisclass}. Each period are analysed separately, but we then combine results
in order to show the development of the social patterning of mortality
during the studied period.

@@

The important explanatory variable, {\sc Hisclass}, is included in the
proportional hazards models as a \emph{stratification} variable. The main
reason for this choice is that the effect of {\sc Hisclass} on survival is
non-proportional, that is, it varies with age. With the reduced data set
and Poisson regression, this is simply achieved by introducing an
interaction term between {\tt age} and {\tt hisclass}. Then, the main
results are presented \emph{graphically}. Tables with results from the different models are presented at the end of the article, but are not commented on in detail. The comments on the results are primarily based on the figures and focuses on the central variable in our study -- {\sc Hisclass}. As can be seen from the tables, most variables in the model are significant at 5 per cent level.

@@

The cumulative
hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
main target in our analysis: These numbers are used in the process of
graphically illustrating the development over time.

@@

Performing the same types of analyses as for all-cause mortality, we start by looking at cumulative hazards by {\sc Hisclass} at age 65, given survival to age 40.

@@

In this section we focus on cardiovascular mortality among those that mostly have left the workforce, the age group 65--89 years. We use as a summary measure the value of the cumulative hazards at
duration 25 (age 90), starting from age 65.
